Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools

### 2024-06-23 paper-r1.qmd
R2 version

---
title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
# subtitle: A case study of Ireland
bibliography: references.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
execute: 
  echo: false
  message: false
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
  # html: default
  # Uncomment for formatted version
  elsevier-pdf:
    linestretch: 2
    keep-tex: true
    journal:
      name: European Transport Research Review
      formatting: preprint
      model: 3p
      cite-style: super
# # # #   # pdf: default
# # Uncomment the following line to produce the cover page
# author:
#   - name: Robin Lovelace
#     email: r.lovelace@leeds.ac.uk
#     corresponding: true
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Joey Talbot
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Eugeni Vidal-Tortosa
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Hussein Mahfouz
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Elaine Brick
#     affiliations:
#       - name: AECOM
#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
#   - name: Peter Wright
#     affiliations:
#       - name: AECOM
#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
#   - name: "Gary O'Toole"
#     affiliations:
#       - name: AECOM
#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
#   - name: Dan Brennan
#     affiliations:
#       - name: Transport Infrastructure Ireland
#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
#   - name: Suzanne Meade
#     affiliations:
#       - name: Transport Infrastructure Ireland
#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  This paper describes an approach for developing strategic cycle network planning tools. Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface that has been tested and used by practitioners. Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions. The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions. CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level. Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks. Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning. The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
  - Transport Planning
  - Collaborative Planning
date: last-modified
---

@@

```{r}
#| include: false
#| eval: false
remotes::install_dev("tmap")
library(tmap)
# Save title page (1st page of paper.pdf):
system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
# Combine cv.pdf with visions.pdf:
system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
# latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
system("latexdiff paper-r1.tex paper.tex > tracked-changes-r2.tex")
system("pdflatex tracked-changes-r2.tex")
browseURL("tracked-changes-r2.pdf")
```

@@

Tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
These tools can be considered as a specific application of the concept of planning support systems (PSS), which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.

@@

```{r}
#| label: tbl-tools
#| tbl-cap: "Summary of existing tools for estimating cycling potential."
# # For PDF:
knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
  kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
  kableExtra::column_spec(1, width = "13em") |>
  kableExtra::column_spec(2, width = "15em") |>
  kableExtra::column_spec(3, width = "12em") |>
  kableExtra::column_spec(4, width = "5em")
# # For HTML:
# knitr::kable(tools)
```

@@

Other tools that generate areal data to prioritize investment include a study that generated a ['bikeability index' for the Metro Vancouver region [@winters_mapping_2013], later renamed Bike Score and available for over 160 US and Canadian cities [@winters_bike_2016] -- its methodology is [openly accessible](https://www.walkscore.com/bike-score-methodology.shtml); a]{style="color: blue;"} 'prioritization index' for Montreal, Canada [@larsen_build_2013]; a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil [@zhang_prioritizing_2014]; and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).

A more recent areal-based approach is the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
These indicators are calculated considering topography, road hierarchy, and average congestion and presented on a scale from 5 to 10.
They are then combined into an overall score, weighted by their impact on cycling according to the literature.
The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].
[The GPC tool has also been used to investigate the combination of factors contributing to high levels cycling potential [@lopes_impact_2021]. Another recent areal-based approach is the Relative Equity Impact of Bicycle Planning (TIRE) [@cunha_assessing_2023] tool. The TIRE tool assesses the impact of cycling network allocation on the accessibility levels of different socioeconomic groups.]{style="color: blue;"}

Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace2016].
The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.

@@

[Another route-based tool is The Cycling Routing Algorithm for Network Connectivity (CRANC) [@gehrke_cycling-focused_2020]. The CRANC tool evaluates the benefits of new bicycle facilities for the accessibility of various populations and neighbourhoods.]{style="color: blue;"}

@@

```{r}
#| label: tbl-trip-purposes
#| tbl-cap: "Trip purposes used in the CRUSE Tool."
# For PDF:
# knitr::kable(trip_purposes, booktabs = TRUE) |>
#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
#   kableExtra::column_spec(1, width = "5em") |>
#   kableExtra::column_spec(2, width = "12em") |>
#   kableExtra::column_spec(3, width = "15em") |>
#   kableExtra::column_spec(4, width = "15em")
# For HTML:
knitr::kable(trip_purposes)
```

@@

::: -->
```

## [Stakeholder engagement]{style="color: blue;"} {#sec-stakeholder-engagement}

[The CRUSE Tool was developed for, and in close collaboration with, practitioners in Ireland.
Although the overall approach of the project was quantitative, stakeholder feedback and qualitative data was deemed essential to ensure the tool was fit for purpose, mitigating against the risk that the project would be 'technocentric' and a 'problem looking for a solution'.
From the outset of the project, we aimed to follow @coppola2022's advocacy for a "human-centric approach in which personalization, purpose, ethics and inclusivity are key", rather than an 'off-the-shelf' product that failed to understand, let alone meet, local needs.
As outlined by @coppola2022, the 'digital transformation' taking place in transport planning (and many other sectors) has the potential to increase existing inequalities [@coppola2022].
A wide range of methods and measure can be taken to ensure that digital tools are accessible to their recipients.]{style="color: blue;"}

<!-- Details from final workshop:

Can you please join me and my colleagues from TII and ITS Leeds for 1 hour on Thursday the 7th December at 12:00 – 13:00 for a workshop and demonstration of the open source CRUSE tool) which is used primarily to highlight the cycle friendliness of roads in Ireland, avail of cycle scenario/trip data and highlight areas to target for active travel investment. This tool is free to use for transport planners or member of the general public and can be found here: CRUSE . We have now added recreational cycling trips to the CRUSE tool and would welcome the chance to present this work and ask for feedback.

Please see below a draft agenda for the workshop till 1pm:

-   Introduction (5 min)
-   Changes to CRUSE: extension work (15 min)
-   Testing new website (10 min)
-   Feedback, Q&A + discussion (30 min)
    - General feedback on new site
    - Feedback on the recreational outputs
    - Feedback on the tourism network -->

<!-- – What are main issues you face in
understanding infrastructure?, e.g.
– What is happening at the moment?
– Where to invest in the future?
– What will change?
– How will planning vary depending on trip purpose?
– What sort of scenarios interested in testing?
– How useful is having different networks for
planning infrastructure and/or more detail,
e.g. there could be an option for variation in
access for professionals and public (the latter
remaining simple) -->

[The stakeholder engagement was carried out through workshops with stakeholders in case study area, an approach that is documented with reference to participatory development of digital tools for supporting energy transitions [@hewitt2020].
We held a total of three workshops over the course of the project and one open-ended live demo during a 2-day event in Sligo in which we collected feedback from dozens of people.
Other than the live demo, all of the workshops took place online.]{style="color: blue;"}

[The first two workshops took place in May 2022, during which the tool was introduced to key staff involved in cycle network planning in Kildare and Limerick counties.
The county-specific approach allowed the workshop to focus on specific areas and key corridors.
The local transport planners were asked to provide feedback in a semi-structured interview approach, with questions on the evidence they currently have access, how they may be able to use the tool, and requests for new features.
The third and final workshop took place on the 7^th^ December 2023 and included descriptions of recent changes to the tool, with an aim being to get feedback on the new recreational and cycle touring evidence, with representatives from county-level transport planning bodies, Sport Ireland, and Waterways Ireland.]{style="color: blue;"}

[The open-ended live demo was undertaken at a stall at TII’s National Roads and Greenways Conference in Sligo in September 2022, where much of the feedback was verbal and not collected.
Much of this feedback from the workshops and live demo outlined above was in the form of ad-hoc comments and suggestions and not all of the qualitative data received was recorded in a structured way, a limitation that we refer back to in @sec-discussion.]{style="color: blue;"}

# Results {#sec-results}

@@

## Stakeholder feedback {#sec-qualitative}

@@

[A final limitation of the approach relates to *how* it was developed.
As described in @sec-stakeholder-engagement, the development of the tool was participatory, but only insofar as people who attended our workshops could provide feedback.
The people who did provide feedback were deliberately chosen as the target audience.
In future work we would like to collect feedback from a wider range of people, including campaigners, service providers (e.g. staff involved in travel planning for schools and new developments), and people who not involved in the planning process.]{style="color: blue;"}

>>>>

---
title: "Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools"
# subtitle: A case study of Ireland
bibliography: references.bib
csl: apa.csl
editor:
  markdown:
    wrap: sentence
execute: 
  echo: false
  message: false
number-sections: true
latex-clean: false
pdf-engine: pdflatex
knitr:
  opts_chunk:
    fig.path: images/
format: 
 elsevier-pdf:
   linestretch: 2
   keep-tex: true
   journal:
     name: European Transport Research Review
     formatting: preprint
     model: 3p
     cite-style: super
# # # #   # pdf: default

# # Uncomment the following line to produce the cover page
# author:
#   - name: Robin Lovelace
#     email: r.lovelace@leeds.ac.uk
#     corresponding: true
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Joey Talbot
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Eugeni Vidal-Tortosa
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Hussein Mahfouz
#     affiliations:
#       - name: University of Leeds
#         address: Institute for Transport Studies, Leeds, LS2 9JT, United Kingdom
#   - name: Elaine Brick
#     affiliations:
#       - name: AECOM
#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
#   - name: Peter Wright
#     affiliations:
#       - name: AECOM
#         address: Winslade House, Winslade Park, Manor Drive, Clyst St Mary, Exeter, EX5 1FY, UK
#   - name: "Gary O'Toole"
#     affiliations:
#       - name: AECOM
#         address: Unit 6, Galway Technology Park, Parkmore, Galway, Ireland
#   - name: Dan Brennan
#     affiliations:
#       - name: Transport Infrastructure Ireland
#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
#   - name: Suzanne Meade
#     affiliations:
#       - name: Transport Infrastructure Ireland
#         address: Parkgate Business Centre, Parkgate Street, Dublin 8, D08 DK10, Ireland
abstract: |
  This paper describes an approach for developing strategic cycle network planning tools. Based on our experience developing and deploying the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we outline the underlying methods, including disaggregation of origin-destination data, incorporation of additional trip purposes, routing, scenario generation, and development of an intuitive user interface that has been tested and used by practitioners. Commissioned and used by Transport Infrastructure Ireland, CRUSE provides estimates of current and potential future cycling levels under 'snapshot' scenarios to inform investment decisions. The publicly available results at https://cruse.bike/ enable planners, engineers, and other stakeholders to make more evidence-based decisions. CRUSE goes beyond previous work by: modeling networks at high spatial resolution; simulating multiple trip purposes (social, shopping, personal utility, recreational, and cycle touring) in addition to using official origin-destination data; and providing estimates of 'quietness' (a proxy for cyclist comfort and route preference) at the route segment level. Three network types --- 'Fastest', 'Balanced', and 'Quietest' --- help plan both arterial and residential cycle networks. Workshops and unstructured interviews with stakeholders show that the tool is already being used in practice to inform urban, inter-urban, and rural network planning. The approach is flexible and open-source, allowing the underlying ideas and code to be adapted, supporting effective cycling policies and interventions internationally.
keywords: 
  - Cycling
  - Open-source
  - Road Safety
  - Active Travel
  - Transport Planning
  - Collaborative Planning
date: last-modified
---

@@

```{r}
#| include: false
#| eval: false
remotes::install_dev("tmap")
library(tmap)
# Save title page (1st page of paper.pdf):
system("pdftk _site/paper.pdf cat 1 output _site/title_page.pdf")
# Combine cv.pdf with visions.pdf:
system("pdftk _site/cv.pdf _site/visions.pdf cat output _site/cv_and_visions.pdf")
# latexdiffr::latexdiff("paper-v1.qmd", "paper.qmd")
system("latexdiff paper-v1.tex paper.tex > tracked-changes.tex")
system("pdflatex tracked-changes.tex")
browseURL("tracked-changes.pdf")
```

@@

Tools for estimating geographical distribution of cycling potential have advanced substantially in recent years.
Developed by researchers and developers in academic, public and private sectors, they have evolved from simple area-based static models tailored to specific regions to more complex and (in some cases) more generalizable tools.
These tools can be considered as a specific application of the concept of planning support systems (PSS) , which are designed to help planners and decision-makers to make more informed decisions [@geertman2009].
Outputs include maps with results available at several levels of analysis with levels of availability ranging from being only available to researchers and in static maps to open access mapping systems available to the public.

@@

```{r}
#| label: tbl-tools
# # For PDF:
# knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.", booktabs = TRUE) |>
#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE) |>
#   kableExtra::column_spec(1, width = "13em") |>
#   kableExtra::column_spec(2, width = "15em") |>
#   kableExtra::column_spec(3, width = "12em") |>
#   kableExtra::column_spec(4, width = "5em")
# For HTML:
knitr::kable(tools, caption = "Summary of existing tools for estimating cycling potential.")
```

@@

Other tools that generate areal data to prioritize investment include a study that generated a 'prioritization index' for Montreal, Canada [@larsen_build_2013], a 'usage intensity index' based on stated preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool (CPT) for Scotland [@phillips_development_2017].
The CPT is composed of the base environmental module — based on eight weighted factors (population density, hilliness, physical barriers, access to services, existing cycling mode share, distance to work and school, and road speed) — and the quality of service module, which evaluates each area of interest based on eight factors (surface condition, adjacent cyclists, comfort factor, conflict, distance between junctions, slope, access to services, and origin/destination).

A more recent areal-based approach is the designed the Gross Potential for Cycling (GPC) tool to prioritize areas for cycling infrastructure and other cycling measures [@silva_gross_2021].
The GPC uses two sets of indicators: population-based indicators (age, potential demand density, employment density and motorization rate) and area-based indicators (accessibility, education, public transport, connectivity, land use mix and relative performance).
These indicators are calculated considering topography, road hierarchy, and average congestion and presented on a scale from 5 to 10.
They are then combined into an overall score, weighted by their impact on cycling according to the literature.
The ranking of areas by the GPC are [publicly available](https://boost.up.pt/en/ferramentas/gpc) and its practical value has been assessed through workshops [@silva_revealing_2022].

Route-based tools emerged with increasing availability of origin-destination data, routing engines that can assign trips to networks, and improvements in computer hardware and software needed to generate route networks.
A prominent example is the the Propensity to Cycle Tool (PCT), first developed to estimate current and future levels of cycling at desire line, zone, route, and route network levels for case study cities [@lovelace_mapping_2016].
The approach was scaled-up to estimate estimate the potential benefits of uptake at zone and desire line levels nationally, and launched as a publicly available web application in 2017 [@lovelace2017].
Extensions of the PCT approach have included estimation of benefits at the individual level [@woodcock2018], addition of travel to school network [@goodman2019], and improved modeling of impacts on health, environmental and distributional outcomes [@woodcock2021].
Initially developed just for England, the PCT was extended to cover all of Wales (for commuter data only) in 2018.

@@

@@

```{r}
#| label: tbl-trip-purposes
#| tbl-cap: "Trip purposes used in the CRUSE Tool."
# For PDF:
# knitr::kable(trip_purposes, booktabs = TRUE) |>
#   kableExtra::kable_styling(latex_options = "striped", full_width = FALSE, font_size = 8) |>
#   kableExtra::column_spec(1, width = "5em") |>
#   kableExtra::column_spec(2, width = "12em") |>
#   kableExtra::column_spec(3, width = "15em") |>
#   kableExtra::column_spec(4, width = "15em")
# For HTML:
knitr::kable(trip_purposes, caption = "Trip purposes used in the CRUSE Tool.")
```

@@

::: -->
```
# Results {#sec-results}

@@

## Stakeholder engagement and feedback {#sec-qualitative}

@@
